Taking Linear Logic Apart

### 2018-04-26 doc/tlla2018/main.tex
Minor corrections

Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \begin{array}{lcllcllcllcl}
         \ty{(A \tens B)^\bot} &=& \ty{A^\bot \parr B^\bot}
      &  \ty{\one^\bot}        &=& \ty{\bot}
      &  \ty{(A \parr B)^\bot} &=& \ty{A^\bot \tens B^\bot}
      &  \ty{\bot^\bot}        &=& \ty{\one}
      \\ \ty{(A \plus B)^\bot} &=& \ty{A^\bot \with B^\bot}
      &  \ty{\nil^\bot}        &=& \ty{\top}
      &  \ty{(A \with B)^\bot} &=& \ty{A^\bot \plus B^\bot}
      &  \ty{\top^\bot}        &=& \ty{\nil}
    \end{array}
  \]
\end{definition}
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements can be constructed using the following inference rules.
  \\[0.5\baselineskip]
  Structural Rules.
  \begin{center} \cpInfAx     \cpInfCut   \end{center}\vspace*{0.5\baselineskip}

  Logical Rules.
  \begin{center} \cpInfTens   \cpInfParr  \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfOne    \cpInfBot   \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfPlus1  \cpInfPlus2 \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfWith               \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfNil    \cpInfTop   \end{center}\vspace*{0.5\baselineskip}
\end{definition}
\cp enjoys subject reduction, progress, and termination~\cite{wadler2012}.
\begin{lemma}[Preservation for $\equiv$]\label{lem:cp-preservation-equiv}
  If $\tm{P}\equiv\tm{Q}$, then $\seq[P]{\Gamma}$ iff $\seq[Q]{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}
\begin{theorem}[Preservation]\label{thm:cp-preservation}
  If $\seq[P]{\Gamma}$ and $\reducesto{P}{Q}$, then $\seq[Q]{\Gamma}$.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reducesto{P}{Q}$.
\end{proof}
\begin{theorem}[Progress]\label{thm:cp-progress}
  If $\seq[P]{\Gamma}$, then there exists a $\tm{Q}$ such that
  $\tm{P}\Longrightarrow^\star\tm{Q}$ and $\tm{Q}$ is not a cut.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\seq[P]{\Gamma}$.
  If the last rule is \textsc{Cut}, there are four cases:
  \begin{enumerate*}[label={\alph*)}]
  \item
    if either side of the cut is an axiom, we apply \cpRedAxCut1;
  \item
    if either side of the cut is itself a cut, we recursively eliminate the cut;
  \item
    if both sides are logical rules acting on the cut formula, we apply
    one of the $\beta$-rules;
  \item
    otherwise, at least one side is a logical rule acting on a formula other
    than the cut formula, in which case we apply one of the $\kappa$-rules.
  \end{enumerate*}
\end{proof}

@@

\begin{figure*}[!htb]
  Structural Rules.
  \begin{center} \hccpInfAx        \hccpInfCycle \end{center}
  \begin{center} \hccpInfMix       \hccpInfHalt  \end{center}

  Logical Rules.
  \begin{center} \hccpInfBoundTens \hccpInfParr  \end{center}
  \begin{center} \hccpInfOne       \hccpInfBot   \end{center}
  \begin{center} \hccpInfPlus1     \hccpInfPlus2 \end{center}
  \begin{center} \hccpInfWith                    \end{center}
  \begin{center} \hccpInfNil       \hccpInfTop   \end{center}

  \centering
  (Where each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.)

  \caption{Hypersequent Classical Processes (\hcp)}
  \label{fig:hccp}
\end{figure*}

@@

\begin{itemize}
\item Every \cp process is an \hcp process.
\item \hcp processes represent multiple entangled \cp processes.
\item \hcp supports the same protocols as \cp.
\end{itemize}

>>>>

Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \begin{array}{lcllcllcllcl}
         \ty{(A \tens B)^\bot} &=& \ty{A^\bot \parr B^\bot}
      &  \ty{\one^\bot}        &=& \ty{\bot}
      &  \ty{(A \parr B)^\bot} &=& \ty{A^\bot \tens B^\bot}
      &  \ty{\bot^\bot}        &=& \ty{\one}
      \\ \ty{(A \plus B)^\bot} &=& \ty{A^\bot \with B^\bot}
      &  \ty{\nil^\bot}        &=& \ty{\top}
      &  \ty{(A \with B)^\bot} &=& \ty{A^\bot \plus B^\bot}
      &  \ty{\top^\bot}        &=& \ty{\nil}
    \end{array}
  \]
\end{definition}
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements can be constructed using the following inference rules.
  \\[0.5\baselineskip]
  Structural Rules.
  \begin{center}\cpInfAx     \cpInfCut  \end{center}\vspace*{0.5\baselineskip}

  Logical Rules.
  \begin{center}\cpInfTens   \cpInfParr \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\cpInfOne    \cpInfBot  \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\cpInfPlus1  \cpInfPlus2\end{center}\vspace*{0.5\baselineskip}
  \begin{center}\cpInfWith              \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\cpInfNil    \cpInfTop  \end{center}\vspace*{0.5\baselineskip}
\end{definition}
\cp enjoys subject reduction, progress, and termination~\cite{wadler2012}.
\begin{lemma}[Preservation for $\equiv$]\label{lem:cp-preservation-equiv}
  If $\tm{P}\equiv\tm{Q}$, then $\seq[P]{\Gamma}$ iff $\seq[Q]{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}
\begin{theorem}[Preservation]\label{thm:cp-preservation}
  If $\seq[P]{\Gamma}$ and $\reducesto{P}{Q}$, then $\seq[Q]{\Gamma}$.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reducesto{P}{Q}$.
\end{proof}
\begin{theorem}[Progress]\label{thm:cp-progress}
  If $\seq[P]{\Gamma}$, then there exists a $\tm{Q}$ such that
  $\tm{P}\Longrightarrow^\star\tm{Q}$ and $\tm{Q}$ is not a cut.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\seq[P]{\Gamma}$.
  If the last rule is \textsc{Cut}, there are four cases:
  \begin{enumerate*}[label={\alph*)}]
  \item
    if either side of the cut is an axiom, we apply \cpRedAxCut1;
  \item
    if either side of the cut is itself a cut, we recursively eliminate the cut;
  \item
    if both sides are logical rules acting on the cut formula, we apply
    one of the $\beta$-rules;
  \item
    otherwise, at least one side is a logical rule acting on a formula other
    than the cut formula, in which case we apply one of the $\kappa$-rules.
  \end{enumerate*}
\end{proof}

@@

\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  interleaved processes, each of which communicates according to its own
  protocol $\Gamma_i$. 
  Typing judgements can be constructed using the following inference rules.
  \\[0.5\baselineskip]
  Structural Rules.
  \begin{center}\hccpInfAx        \hccpInfCycle\end{center}\vspace*{0.5\baselineskip}
  \begin{center}\hccpInfMix       \hccpInfHalt \end{center}\vspace*{0.5\baselineskip}

  Logical Rules.
  \begin{center}\hccpInfBoundTens \hccpInfParr \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\hccpInfOne       \hccpInfBot  \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\hccpInfPlus1     \hccpInfPlus2\end{center}\vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith                   \end{center}\vspace*{0.5\baselineskip}
  \begin{center}\hccpInfNil       \hccpInfTop  \end{center}\vspace*{0.5\baselineskip}

  Where each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}

@@

\begin{itemize}
\item Not every \hcp process is a \cp process.
\item Every \cp process is an \hcp process.
\item \hcp processes represent multiple entangled \cp processes.
\item \hcp supports the same protocols as \cp.
\end{itemize}
